---
title: "Introduction to SIACS: Simulation of Indoor Air Chemistry and Surfaces"
author: "SIACS Development Team"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Introduction to SIACS}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment  = "#>",
  eval     = FALSE   # set FALSE so examples don't run during CRAN check
)
```

## Overview

**SIACS** (Simulation of Indoor Air Chemistry and Surfaces) is an ODE-based
single-zone box model for simulating indoor air quality. It solves the
mass-balance equations for all tracked chemical species simultaneously,
accounting for:

- Gas-phase photochemistry (SAPRC99 or SAPRC07T, ~211 reactions)
- Ventilation (infiltration, balanced, unbalanced, and natural ventilation)
- Surface deposition (time-varying, activity-dependent)
- Indoor emissions from occupant activities
- Indoor light environment from outdoor solar radiation and artificial lights
- Aerosol mass (PM₂.₅, PM₂.₅₋₁₀) and particle number concentrations

Optional post-processing analyses include time derivatives (dy/dt), mass
balance component separation, first-order sensitivity coefficients, and
uncertainty propagation from input uncertainties.

---

## Installation

```{r install}
# From local source:
install.packages("path/to/SIACS_0.2.0.tar.gz", repos = NULL, type = "source")

# Load the package:
library(SIACS)
```

---

## Usage Mode 1: Interactive GUI

The easiest way to get started is the Shiny GUI:

```{r gui}
library(SIACS)
runSIACSApp()
```

This opens a browser window with two input interfaces:

**Wizard Mode** (recommended for new users)  
A step-by-step guided workflow:

1. **Location & Time** — set latitude, longitude, start date/time, duration
2. **Building** — floor area, room height, area-to-volume ratio
3. **Windows** — orientation, area fraction, glass type, obstruction
4. **Artificial Lights** — lamp type, geometry, schedule
5. **Physical Environment** — indoor/outdoor temperature, humidity, pressure,
   ventilation rates (as time schedules)
6. **Outdoor Concentrations** — key species concentrations (ppm, time series)
7. **Emissions & Activities** — select activity sources and schedules
8. **Outputs** — choose output file names and optional analyses

Default values are pre-loaded from the bundled example Atlanta scenario.

**Advanced Mode**  
Tab-based interface for uploading complete CSV/XLSX input files directly.
Gives full control over all parameters.

---

## Usage Mode 2: Batch Scripting

For multiple scenarios or integration into automated workflows, use
`SIACS.batch()`.

### Step 1: Prepare the input list

Each element of `input_data_list` is a list of data frames, one per
simulation instance. The example input files bundled with the package
(accessible via `system.file("extdata", "Input", package = "SIACS")`)
show the required format for each input type.

```{r batch-inputs}
library(SIACS)
library(readxl)

input_dir <- system.file("extdata", "Input", package = "SIACS")

# Read example inputs for one instance:
time_data   <- read.csv(file.path(input_dir, "Time.csv"))
box_data    <- read.csv(file.path(input_dir, "BoxData - Atlanta.csv"))
phys_env    <- read.csv(file.path(input_dir,
                 "PhysicalEnvironmentData_CMAQ Atlanta_48hrs.csv"))
outdoor     <- read.csv(file.path(input_dir,
                 "Outdoor Concentrations - Atlanta July CMAQ.csv"))
emissions   <- read.csv(file.path(input_dir, "EmissionProfiles.csv"))
activities  <- read.csv(file.path(input_dir, "Activities.csv"))
deposition  <- read.csv(file.path(input_dir, "Vd&P-Carslaw 2012.csv"))
initial_val <- read.csv(file.path(input_dir,
                 "InitialIndoorConcentrations.csv"))
indoor_light <- read_excel(file.path(input_dir, "IndoorLightAtlanta.xlsx"),
                            sheet = "J_values")
```

### Step 2: Assemble the input data list

```{r batch-assemble}
input_data_list <- list(
  Time                 = list(time_data),
  BoxData              = list(box_data),
  PhysicalEnvironment  = list(phys_env),
  OutdoorConcentrations = list(outdoor),
  EmissionProfiles     = list(emissions),
  Activities           = list(activities),
  DepositionV          = list(deposition),
  InitialValues        = list(initial_val),
  IndoorLight          = list(
    list(J_values    = as.data.frame(indoor_light),
         EnergyFlux  = data.frame(Time = ..., EnergyFlux = ...))
  )
)
```

### Step 3: Define outputs

```{r batch-outputs}
output_list <- list(
  list(
    OutputTable                  = "./Output/Results",
    OutputBasicChart             = "./Output/Plot",
    OutputTimeDerivatives        = "./Output/Derivatives",
    OutputMassBalanceComponents  = "./Output/MassBalance",
    OutputSensitivity            = "None",
    OutputUncertainty            = "None"
  )
)
```

### Step 4: Run

```{r batch-run}
results <- SIACS.batch(
  input_data_list = input_data_list,
  instances       = 1,
  OutputList      = output_list,
  mechanism       = "SAPRC99",
  chemistry       = TRUE,
  use_parallel    = FALSE   # use TRUE for multiple instances
)
```

### Step 5: Inspect results

```{r batch-results}
# Main output data frame:
out <- results[[1]]$Simulation1$alldata

# View key species over time:
head(out[, c("time", "O3", "NO2", "HCHO", "OH", "PM25")])

# Time is in minutes; concentrations in ppm (gases) or µg/m³ (PM)
```

---

## Input File Formats

### Time (`Time.csv`)
| Column | Description | Units |
|---|---|---|
| `StartTimeYear` | Calendar year | integer |
| `StartTimeMonth` | Calendar month | 1–12 |
| `StartTimeDay` | Calendar day | 1–31 |
| `StartTime` | Local clock time (HH:MM) | character |
| `StartTimeStandard` | Timezone code (e.g., EST) | character |
| `RelativeStartTime` | Minutes from midnight | minutes |
| `TimeStep` | Output time step | minutes |
| `Duration` | Simulation length | hours |

### Physical Environment (`PhysicalEnvironmentData.csv`)
First row contains relative uncertainty values (fractions). Data rows
contain time series with columns: `Time` (minutes), `Ti` (indoor temp, K),
`To` (outdoor temp, K), `RH` (relative humidity, 0–1), `BP` (barometric
pressure, Pa), `QBal` (balanced ventilation, m³/s), `QUnbal` (unbalanced,
m³/s), `QNat` (natural, m³/s), `QFilter` (filter recirculation, m³/s),
`Wind` (wind speed, m/s).

### Outdoor Concentrations (`OutdoorConcentrations.csv`)
First row: relative uncertainty fractions. Subsequent rows: time (minutes)
and species concentrations in ppm. Columns are named by species code
(e.g., `O3`, `NO2`, `NO`, `CO`, `SO2`, `HCHO`). Missing species default
to zero.

### Emission Profiles (`EmissionProfiles.csv`)
First row: relative uncertainty fractions. Subsequent rows: one row per
emission source profile, with `ProfileName` column and species columns
giving emission rates in g/min per unit of source activity.

---

## Chemical Mechanisms

SIACS implements two SAPRC-based gas-phase chemistry mechanisms:

| Mechanism | Reference | Reactions |
|---|---|---|
| `SAPRC99` | Carter (2000) | 211 |
| `SAPRC07T` | Carter (2010) | 211 |

Both mechanisms include photolysis reactions whose rates are computed from
the indoor light flux (J-values) calculated by the TUV model integration.
Aggregated diagnostic species are also calculated and included in output:
ALK (total alkanes), ARO (aromatics), OLE (alkenes), NOy (total reactive
nitrogen), RO2 (peroxy radicals), HOx (OH + HO₂ + RO₂).

---

## Output Variables

The main output data frame (`alldata`) contains columns for:

- `time` — simulation time (minutes)
- Indoor concentrations for all species (ppm for gases; µg/m³ for PM)
- Outdoor concentrations for all species (suffix `.O`)  
- Indoor emission source strengths (g/min for gases; µg/min for PM; suffix `.S`)
- Physical environment: `Ti`, `To` (°C), `a` (air exchange rate, h⁻¹),
  `LightFlux` (mW/cm²), `RH` (%), `BP` (Pa), `Mair.in` (molecules/cm³),
  `QFilter` (m³/s)
- Aggregated species: ALK, ARO, OLE, NOy, RO2, HOx (and outdoor counterparts)

---

## Citation

```{r citation}
citation("SIACS")
```

---

## References

Carter, W. P. L. (2000). Documentation of the SAPRC-99 chemical mechanism
for VOC reactivity assessment. Final Report to the California Air Resources
Board.

Carter, W. P. L. (2010). Development of the SAPRC-07 chemical mechanism.
*Atmospheric Environment*, 44(39), 5324–5335.

Weschler, C. J., & Nazaroff, W. W. (2023). Human skin oil: a major ozone
reactant indoors. *Environmental Science: Atmospheres*, 3(4), 640–661.
